Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-17 writeup/pnas-word-order-si.tex
SI

TODO figure out why

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}} is formalized using \emph{obl}, which covers not only PPs and standards of comparison, but also adjunct NPs.
		In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{nmod} agrees with that annotated for obliques in all 18 languages for which WALS has an entry.

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}} is formalized using \emph{xcomp}, which covers other control verbs, not just verbs of volition (`want').

@@

		For each language, we extracted all lemmas of words heading an \emph{xcomp} dependency, occurring at least 2 times.
In 39 languages, all extracted words had the same dominant order.
Additionally, in four Germanic languages (Afrikaans, Danish, Dutch, and German), the verb of volition (Afrikaans \emph{wil}, Danish \emph{ville}, Dutch \emph{willen}, German \emph{wollen}) is mostly annotated with the \emph{aux} relation due to UD annotation guidelines, but in all languages, its dominant order (verb of volition before its complement) agrees with the dominant order of the \emph{xcomp} dependency (head-initial).
		In three historical languages (Ancient Greek, Latin, and Old Church Slavonic), verbs of volition agree with the dominant order of \emph{xcomp}, while several other verbs that do not indicate volition show opposite dominant order.
Finally, in Gothic, the verb of volition (\emph{wiljan}) has opposite order, resulting in an apparent violation of Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption[Predictability and Parseability]{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\footnotemark All data are $z$-scored.}\label{fig:pareto-per-lang}
\end{figure}

@@

>>>>

The exceptions mainly arise because some languages have multiple common word orders for relativization:
Hindi uses correlatives that can precede or follow the coreferent noun \citep[3.1.3]{montaut2004grammar} and relatives following the noun \citep[4.3]{montaut2004grammar}.
Estonian and Finnish have finite relative clauses following nouns (\citep[p. 176]{karlsson2013finnish}, \citep[p. 256]{tauli1983standard}) and nonfinite participial modifiers preceding it \citep[Chapter 18]{karlsson2013finnish}.

		Finally, in Tamil, the divergence is caused by the treebank annotation convention for Tamil, where the label \emph{acl} is used to mark certain elements of compounds whose positioning differs from the participial phrases that correspond most closely to relative clauses of other languages.\footnote{In the original HamleDT \citep{zeman2012hamledt, Ramasamy:2011:TDP:1964799.1964808} version of the Tamil treebank, these relations were labeled as CC, marking compounds (\url{http://ufal.mff.cuni.cz/~ramasamy/tamiltb/0.1/dependency_annotation.html}). We did not attempt to modify this labeling convention.}

\item Correlation {\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}} is formalized using \emph{obl}, which covers not only PPs and standards of comparison, but also adjunct NPs.
In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{nmod} agrees with that annotated for obliques in all 18 languages for which WALS has an entry.

\item Correlation {\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}} is formalized using \emph{xcomp}, which covers other control verbs, not just verbs of volition (`want').

@@

For each language, we extracted all lemmas of words heading an \emph{xcomp} dependency, occurring at least 2 times.
In 39 languages, all extracted words had the same dominant order.
Additionally, in four Germanic languages (Afrikaans, Danish, Dutch, and German), the verb of volition (Afrikaans \emph{wil}, Danish \emph{ville}, Dutch \emph{willen}, German \emph{wollen}) is mostly annotated with the \emph{aux} relation due to UD annotation guidelines, but in all languages, its dominant order (verb of volition before its complement) agrees with the dominant order of the \emph{xcomp} dependency (head-initial).
In three historical languages (Ancient Greek, Latin, and Old Church Slavonic), verbs of volition agree with the dominant order of \emph{xcomp}, while several other verbs that do not indicate volition show opposite dominant order.
Finally, in Gothic, the verb of volition (\emph{wiljan}) has opposite dominant order, resulting in an apparent violation of Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage.pdf}
\caption[Predictability and Parseability]{Predictability and parseability of 51 languages. Green: random baselines, Red: real grammar, blue: approximate Pareto frontier, computed from the optimized grammars.\footnotemark All data are $z$-scored. TODO adapt for colorblindness.}\label{fig:pareto-per-lang}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-isoCurve.pdf}
\caption{Per-language results. Orange curves connect points estimated to be dominated by the same number of baselines as the real language, obtained by Bayesian inference using a Dirichlet Gaussian Process model.}
\end{figure}
